---
title: "Knitr Project"
author: "A.Rachid"
date: "2024-10-15"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(lattice)
```

# Activity Monitoring Data Analysis

## Loading the data

```{r loading the data}
data=read.csv(file = 'activity.csv')
head(data)
```

## Preprocessing the data

```{r processing the data}
data$date<-as.Date(data$date)
class(data$date)
```

## What is mean total number of steps taken per day?
The sum of the steps taken per day
```{r steps per day,cache=TRUE}
Steps_by_day <- data %>% group_by(date) %>% summarise(steps_h = sum(steps))
head(Steps_by_day)
```
Histogram of the steps taken per day
```{r}
hist(Steps_by_day$steps_h)
```
The mean is :
```{r}
mean(Steps_by_day$steps_h,na.rm=TRUE)
```
The median is :
```{r}
median(Steps_by_day$steps_h,na.rm=TRUE)
```

## What is the average daily activity pattern?
Plotting
```{r plot,cache=TRUE}
avg_step_interval <- aggregate(steps ~ interval, data = data, mean, na.rm = TRUE)
plot(avg_step_interval$interval, avg_step_interval$steps, type = "l", lwd = 2, col = "navy",
     main = "Average Number of Steps Taken", axes = FALSE,
     xlab = "Interval", ylab = "Average number of steps")
axis(1)
axis(2, las = 1)
```
 
The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is as follow :
 
```{r max}
avg_step_interval$interval[which.max(avg_step_interval$steps)]
```

## Imputing missing values
The total number of missing values in the dataset:
```{r}
sum(is.na(data$steps))
```
New data filling the missing values with the following strategy :
Using the mean from each interval to fill in the values missing from a sample
```{r,cache=TRUE}
new_data <- data 
for (i in avg_step_interval$interval) {
    new_data[new_data$interval == i & is.na(new_data$steps),]$steps <- avg_step_interval$steps[avg_step_interval$interval == i]
}
head(new_data)
```

Making a histogram of the total number of steps taken each day, calculating and reporting the mean and median total number of steps taken per day :

```{r}
new_Steps_by_day <- new_data %>% group_by(date) %>% summarise(steps_h = sum(steps))
print(mean(new_Steps_by_day$steps_h))
print(median(new_Steps_by_day$steps_h))
hist(new_Steps_by_day$steps_h)

```

The mean remains consistent with the value obtained in the first part of the assignment, but the median differs slightly, despite being close in value. Imputing missing data with the average for each 5-minute interval introduces more data points equal to the mean, reducing the variability in the distribution. As a result, with many data points matching the mean, the median is more likely to converge toward the mean as well.

## Are there differences in activity patterns between weekdays and weekends?
Factoring the 
```{r}
f <- new_data %>% mutate(weekday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
new_data$weekday<-f$weekday
new_data$weekday <- factor(new_data$weekday, levels = c("weekday", "weekend"))
head(new_data)
```
```{r}
steps <- aggregate(steps ~ interval + weekday, data = new_data, mean)
xyplot(steps ~ interval | weekday, data = steps, type = "l", lwd = 2,
       xlab = "Interval", 
       ylab = "Average number of steps")
```